---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fqr

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/be-green/fqr/branch/main/graph/badge.svg)](https://codecov.io/gh/be-green/fqr?branch=main)
[![R-CMD-check](https://github.com/be-green/fqr/workflows/R-CMD-check/badge.svg)](https://github.com/be-green/fqr/actions)
<!-- badges: end -->

The `fqr` package makes quantile regression fast and scaleable using accelerated
gradient descent. For both big and small problems it is substantially faster than other quantile regression approaches. `fqr` can handle quantile regression problems on the order of 10 million rows and 100 columns in less than a minute, and can exactly match existing implementations on small problems.

While the quantile loss function isn't differentiable, you can get an arbitrarily close smooth approximation by replacing the "check" function with an appropriately tilted least squares approximation for a small neighborhood around the origin. As the size of that window goes to zero, you have your check function back!

The package uses 2 stopping rules to assess convergence: the maximum value of
the gradient vector (for the coefficients of the quantile regression) and the
relative change in the loss function (scaled by the step size).

`fqr` is substantially faster than the `quantreg` package's interior
point methods (e.g. "br" or "sfn"), especially for large problems. The algorithm implemented via the Armadillo library for linear algebra in C++. It also has no dependencies other than base R and (if building from source) a C++ compiler.

## Installation

You can install the fqr package from github by running

```{r, eval = F}
# get remotes if needed:
# install.packages("remotes")

remotes::install_github("be-green/fqr")
```

## Basic Use

The `fqr` package uses the same basic formula interface that `lm` does, with
standard errors calculated based on subsampling.

```{r}
library(fqr)
data(rocks)

fqr(area ~ peri, data = rock, tau = c(0.25, 0.5, 0.75))
```

To turn off standard errors (and just get point predictions), you can set
`se = F`.

```{r}
fqr(area ~ peri, data = rock, se = F, tau = c(0.25, 0.5, 0.75))
```

## Benchmarks

Ok, but _how_ fast is this approach? Let's just take some point estimates and see
how it goes.

### Medium N, Medium P

 But with all of this done, let's compare to some benchmarks from the `sfn` and `pfn` algorithms, which are currently the fastest in the `quantreg` package.
 
```{r}
# simulate some data, 101 x 100,000
p <- 20
n <- 1e6
beta <- rnorm(p + 1)

x <- cbind(1, matrix(rnorm(p * n), ncol = p, nrow = n))
y <- x %*% beta + exp(rnorm(n, sd = 2))

# let's take a look at what this looks like
hist(y)
```

Ok so we have some _very_ skewed data! Perfect for median regression.

```{r cache = F}
# remove the intercept since it's implied in the formula
start = proc.time()
# lower level version that just takes design matrix
fit <- fit_fqr(x, y, tau = 0.5, se = F)
end = proc.time()
end - start
```
I attempted to run the same thing with the `quantreg` package, with the method advised for large datasets, like so: 
```{r eval = F}
# newton interior point method w/ pre-processing
start <- proc.time()
fit_pfn <- quantreg::rq.fit.pfn(x, y, tau = 0.5)
end <- proc.time()
end - start
```

but I killed it after 20 minutes (feel free to try this yourself!).

# Big N, Big P

Let's benchmark with a bigger set of columns.

```{r}
p <- 100
n <- 1e6
beta <- rnorm(p + 1)

x <- cbind(1, matrix(rnorm(p * n), ncol = p, nrow = n))
y <- 10 + x %*% beta + exp(rnorm(n, sd = 2))
```


```{r cache = F}
start = proc.time()
fit <- fit_fqr(x, y, tau = 0.5, se = F)
end = proc.time()
end - start
```

I'm not going to run the quantreg `pfn` algorithm since it was so slow for the last problem.

### Big N, Small P
Let's try a more manageable set of dimensions, with _lots_ of observations.

```{r}
p <- 10
n <- 1e7
beta <- rnorm(p + 1)

x <- cbind(1, matrix(rnorm(p * n), ncol = p, nrow = n))
y <- x %*% beta + exp(rnorm(n, sd= 2))

start = proc.time()
fit <- fit_fqr(X = x, y = y, tau = 0.5, se = F)
end = proc.time()
end - start
```
I attempted to do the comparable thing for the `pfn` algorithm:
```{r, eval = F}
start = proc.time()
fit <- quantreg::rq.fit.pfn(x = x, y = y, tau = 0.5)
end = proc.time()
end - start
```
...but I killed the process after 15 minutes or so.


# Medium-scale Problem

Ok, so we haven't been able to run quantreg on these datasets, let's see how it does with a sort of medium-scale problem. Let's use the same DGP.

```{r}
p <- 10
n <- 1e5
beta <- rnorm(p + 1)

x <- cbind(1, matrix(rnorm(p * n), ncol = p, nrow = n))
y <- x %*% beta + exp(rnorm(n, sd= 2))

start = proc.time()
fit <- fit_fqr(X = x, y = y, tau = 0.5, se = F)
end = proc.time()
end - start
```

```{r, eval = T, message = F}
start = proc.time()
fit_pfn <- quantreg::rq.fit.pfn(x = x, y = y, tau = 0.5)
end = proc.time()
end - start
```

The coefficients match out to the 4th or 5th decimal place:
```{r}
max(abs(fit$coefficients - fit_pfn$coefficients))
```

```{r}
min(abs(fit$coefficients - fit_pfn$coefficients))
```




### Small Problems

It can also be faster for small problems, and with conservative tolerance parameters will come extremely close to the default `quantreg` outputs. Here's an example: 

```{r cache = F}
# simulate some data, 101 x 10,000,000
p <- 3
n <- 10000
beta <- rnorm(p)
x <- cbind(matrix(rnorm(p * n), ncol = p, nrow = n))
y <- 10 + x %*% beta + exp(rnorm(n, sd = 2))

microbenchmark::microbenchmark(
  fqr_fit <- fqr(y ~ ., se = F, beta_tol = 0, check_tol = 0,
                 data = data.frame(y = y, x)),
  br_fit <- quantreg::rq(y ~ ., tau = 0.5, 
                    data = data.frame(y = y, x), method = "br"),
                    times = 100
)
```

The coefficients match out to 4 decimal places:

```{r}
fqr_fit$coefficients - br_fit$coefficients
```

And the check loss is nearly identical:

```{r}
check <- function (x, tau = 0.5) {
  sum(x * (tau - (x < 0)))
}

check(fqr_fit$residuals) -  check(br_fit$residuals)
```

Still, though, the speed gains are most noticeable once N and P are "medium" or larger (e.g. for N < 300, probably just use quantreg).
